---
title: "Getting started with daisyr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with daisyr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(daisyr)
```

## Overview

daisyr drives the [Daisy](https://daisy.ku.dk/) agroecosystem model (a
soil-crop-water-nitrogen simulator) from R for two workflows:

- **Calibration** - fit free parameters against observed data using an
  optimizer (currently `DEoptim`).
- **Sensitivity analysis** - Morris or Sobol-Jansen designs, run externally
  against Daisy ("decoupled" mode).

Both workflows share the same substrate:

1. A **parameters** YAML file describes every free parameter:
   where it lives in a `.dai` template, its bounds, and - for parameters that
   are shape parameters of a piecewise-linear-function (PLF) - how to turn
   them into the `(x y) (x y) ...` sequence Daisy expects.
2. `render_templates()` substitutes the current parameter values into
   `.dai` template file(s), producing the concrete `.dai` file(s) Daisy will
   run.
3. `run_daisy()` invokes the Daisy executable.
4. `read_dlf()` parses Daisy's `.dlf` log output.
5. `compute_generic_objective()` (or a custom `response_fun`) turns that
   output, plus observed data, into a score.

This vignette walks through a worked example exercising every step. It's
based on `examples/calibration/run_calibration.R` in the package source,
which you can run directly for a plain-R-script version of the same
workflow.

```{r daisy-check, include = FALSE}
# Point this at your local Daisy installation. If it isn't found, the
# Daisy-running chunks below are skipped (shown as code, not evaluated) so
# this vignette still builds on machines without Daisy installed - only the
# schema/API-focused sections evaluate unconditionally.
daisy_exe <- "C:/Program Files/Daisy 5.93/bin/daisy.exe"
has_daisy <- file.exists(daisy_exe)
example_dir <- system.file("extdata", "calibration_example", package = "daisyr")

# Run all Daisy output (rendered .dai, Output/) in a scratch directory
# rather than polluting vignettes/ with generated files.
run_dir <- tempfile("daisyr-vignette-")
dir.create(run_dir)
knitr::opts_knit$set(root.dir = run_dir)
```

```{r echo=FALSE, results='asis'}
if (!has_daisy) {
  cat("_Note: this build of the vignette did not find a local Daisy installation, ",
      "so the simulation-running code chunks below are shown but not evaluated._\n")
}
```

## The scenario

The example calibrates two parameters of a single soil horizon - clay
content (`Ap_clay`) and dry bulk density (`Ap_bulk_density`) - against two
monthly soil-water-content readings, using a small `.dai` case adapted from
Daisy's own `sample/test-minimize.dai`. The template and observed data ship
with the package under `inst/extdata/calibration_example/`.

## 1. Parameters

The example ships as `parameters.yaml` under `calibration_example/`.

```{r echo=FALSE, comment=""}
cat(readLines(file.path(example_dir, "parameters.yaml")), sep = "\n")
```

Load and validate it - `validate_param_config()` checks that every declared
`{{name}}` placeholder exists exactly once in its template file, catching
typos before any simulation is run:

```{r config}
config <- read_param_config(file.path(example_dir, "parameters.yaml"))
validate_param_config(config, template_dir = example_dir)
config$parameters
```

`param_config_names()` lists every free parameter (used positionally by
[calibrate_daisy()] and [sa_design()]):

```{r param-names}
param_config_names(config)
```

## 2. Rendering templates

`render_templates()` substitutes the current parameter values into every
`(from_file, to_file)` pair declared in the parameters file:

```{r render, eval = has_daisy}
default_values <- stats::setNames(config$parameters$default, config$parameters$name)
render_templates(config, values = default_values, template_dir = example_dir, output_dir = ".")
```

Placeholders use double braces (`{{name}}`), not single braces - Daisy's own
`.dai` syntax uses single-brace-free `${...}` for internal self-reference
(e.g. inside a `deflog`), so single-brace placeholders would collide with
that.

## 3. Running Daisy and reading output

```{r run, eval = has_daisy}
run_daisy("test-optim.dai", daisy_exe = daisy_exe, working_dir = ".")

sim <- read_dlf("Output/interval_water_content.dlf")
sim <- add_date(sim, "interval_water_content")
sim[, .(Date, hour, tocalibrate)]
```

`read_dlf()` handles both the standard `.dlf` format (a `----` separator
line before the header) and files written with `print_header false`/
`print_dimension false`.

## 4. Defining and evaluating the objective

The Daisy log is hourly; we average to daily so it lines up with the
monthly-snapshot observed dates, then compare with
[compute_generic_objective()] via an [objective_spec()]:

```{r objective, eval = has_daisy}
obs <- data.table::fread(file.path(example_dir, "observed_data.txt"))
obs[, Date := as.Date(Date, format = "%Y/%m/%d")]

daily_sim_reader <- function(path) {
  sim <- add_date(read_dlf(path), "sim")
  sim[, .(tocalibrate = mean(tocalibrate)), by = Date]
}

objective <- objective_spec(
  obs_data   = obs,
  value_map  = list(list(obs_col = "Theta", sim_col = "tocalibrate", label = "Topsoil (0-15cm)")),
  sim_reader = daily_sim_reader,
  metric     = "RMSE"
)

evaluate_objective(objective, "Output/interval_water_content.dlf", return_per_pair = TRUE)
```

`value_map` supports multiple weighted obs/sim column pairs at once (e.g.
several depths or several output variables) - see `?compute_generic_objective`.

## 5. Calibration

`calibrate_daisy()` wraps the render -> run -> score loop in an optimizer
(`DEoptim` by default; also `BOBYQA`, `multi-BOBYQA`, `DDS`, `CMA-ES`),
using each parameter's `min`/`max` as bounds:

```{r calibrate, eval = has_daisy}
fit <- calibrate_daisy(
  config     = config,
  run_file     = "test-optim.dai",
  daisy_exe    = daisy_exe,
  objective    = objective,
  sim_file     = "Output/interval_water_content.dlf",
  working_dir  = ".",
  template_dir = example_dir, output_dir = ".",
  control      = list(itermax = 15, NP = 10)
)

fit$best_params
fit$best_score
```

## 6. Sensitivity analysis

`sa_design()` builds a decoupled Morris (or Sobol-Jansen) design over every
every parameter; `run_sa_design()` evaluates it externally via Daisy;
`complete_sa_analysis()` attaches the responses to compute indices; and
`plot_sa()` draws them (Morris \eqn{\mu^*} vs \eqn{\sigma} by default):

```{r sensitivity, eval = has_daisy}
daily_sim_reader_from_dt <- function(sim_dt) {
  sim_dt <- add_date(sim_dt, "sim")
  sim_dt[, .(tocalibrate = mean(tocalibrate)), by = Date]
}

design <- sa_design(config, method = "morris", r = 4,
                     design = list(type = "oat", levels = 4, grid.jump = 2), seed = 1)

sa_results <- run_sa_design(
  design, config, run_file = "test-optim.dai", daisy_exe = daisy_exe,
  working_dir = ".", template_dir = example_dir, output_dir = ".",
  output_files = c(swc = "Output/interval_water_content.dlf"),
  response_fun = function(outputs) {
    daily <- daily_sim_reader_from_dt(outputs$swc)
    compute_generic_objective(
      obs_data = obs, sim_data = daily,
      value_map = list(list(obs_col = "Theta", sim_col = "tocalibrate")),
      metric = "RMSE"
    )$summary
  }
)

sa <- complete_sa_analysis(design, sa_results$responses)
sa
plot_sa(sa)
```

Note that Morris elementary-effects designs are meant to explore
multi-parameter interactions - `sigma` (the non-linearity/interaction
indicator) isn't meaningful for a single-parameter design, which is why this
example uses two parameters rather than one.

## Piecewise-linear-function (PLF) parameters

Some Daisy parameters are piecewise linear functions - a sequence of `(x y)`
pairs where `x` must be strictly increasing (e.g. LAI vs. development stage).
Calibrating every `y` independently doesn't scale and risks producing
non-physical curves, so daisyr supports generating the whole sequence from a
handful of shape parameters of a monotonic curve family, via a `plf_curves`
entry in the parameters file:

```yaml
parameters:
  - name: LAIvsDS_L
    default: 5
    min: 3
    max: 7
  - name: LAIvsDS_k
    default: 2
    min: 0.5
    max: 5
  - name: LAIvsDS_x0
    default: 0.7
    min: 0.3
    max: 1.2

plf_curves:
  - name: LAIvsDS
    from_file: dk-crop_template.dai
    to_file: dk-crop.dai
    placeholder: LAIvsDS_BLOCK   # a single {{LAIvsDS_BLOCK}} in the template
    x_values: [-0.3, 0.0, 1.0, 2.0]   # fixed knots, never calibrated
    curve: logistic
    params: [LAIvsDS_L, LAIvsDS_k, LAIvsDS_x0]
```

`render_templates()` evaluates the curve at the fixed `x_values` and
substitutes the whole `(x1 y1) (x2 y2) ...` block into the named placeholder.
Built-in curve families ([known_plf_curves()]) are `logistic`, `gompertz`,
`richards`, and `exponential`; [register_plf_curve()] adds custom ones. All
of the built-ins are monotonic for sensible parameter ranges, so x-monotonicity
of the resulting PLF is never at risk - only the shape parameters are
calibrated/varied, never the `x` knots.

```{r plf-curve}
x <- c(-0.3, 0.0, 1.0, 2.0)
y <- render_plf_curve("logistic", x, c(5, 2, 0.7))   # L, k, x0
format_plf_block(x, y)
```

To see what each built-in family looks like before picking one for a
`plf_curves` entry, `plot_plf_curves()` (or `known_plf_curves(plot = TRUE)`)
draws an example of each using representative default shape parameters:

```{r plot-plf-curves, fig.width=6, fig.height=4}
plot_plf_curves()
```

Individual PLF points can still be calibrated directly (without a curve fit)
as a plain scalar parameter with `plf` metadata for documentation - see
`?read_param_config`.

```{r cleanup, include = FALSE}
unlink(run_dir, recursive = TRUE)
```
